---
title: "tidy_data"
output: html_document
---

#libraries
```{r}

library(tidyverse)
library(janitor)
library(here)
library(countrycode)
library(vroom)
library(raster)
library(sf)


```

Capture fisheries production data
```{r}
#Sort the symbols in values out
## "..." = data unavailable (here taken as zero)
## " " = data not separately available (here taken as zero)
## "-" = nil or zero (make zero)
## "0 0" = more than zero but less than half the unit used (taken to be 0.25)
## F = FAO estimate from available sources of information


(capture_prod <- 
   read_csv(here("data", "raw_data", "fisheries_prod_raw.csv")) %>% 
   clean_names() %>% 
   rename(country = country_country,
          species = species_asfis_species,
          fishing_area = fishing_area_fao_major_fishing_area,
          unit = unit_unit) %>% 
   pivot_longer(names_to = "year", values_to = "value", -c(country, species, fishing_area, unit)) %>% 
   filter(!country %in% c("Totals - Number", "Totals - Tonnes - live weight")) %>% 
   mutate(year = gsub("x", "", year) %>% 
            as.numeric,
          flag = case_when(value == "..." ~ "No data",
                           value == " " ~ "Data not separately available",
                           value == "-" ~ "Nil or zero",
                           value == "0 0" ~ "0<x<0.5",
                           grepl(" F", value) ~ "estimate",
                           TRUE ~ "Reported"), 
          value = case_when(value %in% c("...", " ", "-") ~ "0",
                            value == "0 0" ~ "0.25",
                            grepl(" F", value) ~ gsub(" F", "", value),
                            TRUE ~ value) %>% 
            as.numeric)
)

#sorts country coding to deal with non-UTF characters that country code depends on
Encoding(capture_prod$country) <- "latin1" #deals with the non-UTF
capture_prod$country <- iconv(capture_prod$country, "latin1", "UTF-8",sub='')

capture_prod$iso3c <- countrycode(capture_prod$country, origin = "country.name", destination = "iso3c", warn=TRUE)
capture_prod$iso3c <- case_when(capture_prod$country == "Zanzibar" ~ "TZA",
                                TRUE ~ capture_prod$iso3c)


saveRDS(object = capture_prod, file = here("data", "tidy_data", "capture_production_tidy.rds"))


```


Aquaculture production data
```{r}

(aqua_prod_raw <- read_csv(here("data", "raw_data", "aqua-prod-raw.csv")) %>% 
  clean_names() %>% 
  rename(country= country_country,
         species = species_asfis_species,
         area = aquaculture_area_fao_major_fishing_area,
         environment = environment_environment)
)


#sorts country coding to deal with non-UTF characters that country code depends on
Encoding(aqua_prod_raw$country) <- "latin1" #deals with the non-UTF
aqua_prod_raw$country <- iconv(aqua_prod_raw$country, "latin1", "UTF-8",sub='')



aquaculture_prod <- aqua_prod_raw %>% 
  pivot_longer(names_to = "year", values_to = "value", -c(country, species, area, environment, unit)) %>% 
  mutate(iso_3c = countrycode(country, origin = "country.name", destination = "iso3c", warn = TRUE)) %>%
  mutate(iso_3c = case_when(country == "Zanzibar" ~ "TZA",
                            TRUE ~ iso_3c)) %>% 
    mutate(year = gsub("x", "", year) %>% 
              as.numeric,
           flag = case_when(value == "..." ~ "No data",
                           value == " " ~ "Data not separately available",
                           value == "-" ~ "Nil or zero",
                           value == "0 0" ~ "0<x<0.5",
                           grepl(" F", value) ~ "estimate",
                           TRUE ~ "Reported"),
           value = case_when(value %in% c("...", " ", "-") ~ "0",
                           value == "0 0" ~ "0.25",
                           grepl(" F", value) ~ gsub(" F", "", value),
                           TRUE ~ value) %>% 
              as.numeric) %>% 
   filter(country != "Totals")

  
saveRDS(object = aquaculture_prod, file = here("data", "tidy_data", "aquaculture_production_tidy.rds"))

```


Fish food balance data
```{r}

(fish_fbs_raw <- read_csv(here("data", "raw_data", "fish_FBS_1961-2017.csv")) %>% 
   clean_names()
)

#remove every other column from row 6

col_seq <- seq(from = 6, to = 118, by=2)

fish_fbs_raw <- fish_fbs_raw[, -col_seq]

(fish_fbs <- fish_fbs_raw %>% 
      pivot_longer(cols = -c(country_name, faostat_group_name, element_name, unit_name), names_to = "year", values_to = "value") %>%
      mutate(year = gsub("x", "", year))
)


saveRDS(object = fish_fbs, file = here("data", "tidy_data", "fish_fbs_tidy.rds"))

```


Fish trade data - volume, temporal
```{r}

col_seq <- seq(from = 7, to = 93, by=2) #remove every other column from the 7th

seafood_trade_raw <- read_csv(here("data", "raw_data", "fish_trade_volume_1976-2019.csv")) %>% 
   clean_names() %>% 
   select(-col_seq) %>% 
   filter(!reporting_country_name %in% c("Totals - Tonnes – net product weight", "FAO. 2021. Fishery and Aquaculture Statistics. Global Fish Trade - All partners aggregated 1976-2019 (FishstatJ). In: FAO Fisheries and Aquaculture Division [online]. Rome. Updated 2021. www.fao.org/fishery/statistics/software/fishstatj/en"))


seafood_trade_volume <- seafood_trade_raw %>% 
   pivot_longer(names_to = "year", values_to = "value", cols=-c(1:5)) %>% 
   mutate(year = gsub("x", "", year) %>% 
             as.numeric)

saveRDS(object = seafood_trade_volume, file = here("data", "tidy_data", "seafood_trade_volume_product_weight_1976-2019.rds"))
   


```


Fish trade data - value, temporal
```{r}

(seafood_value <- read_csv(here("data", "raw_data", "fish_trade_value_1976-2019.csv")) %>% 
   clean_names() %>% 
    select(-seq(from = 7, to = 93, by =2)) %>% 
 pivot_longer(cols=-c(1:5), names_to = "year", values_to = "value") %>% 
    mutate(year = gsub("x", "", year) %>% 
              as.numeric,
           value = value*1000,
           unit = "USD",
           unit_name = "Value")
)

saveRDS(object = seafood_value, file = here("data", "tidy_data", "seafood_trade_value_product_weight_1976-2019.rds"))

```



Fish trade - volume by trade partner in 2019
```{r}

seafood_trade_volume_2019 <- read_csv(here("data", "raw_data", "fish_trade_volume_by_partner_2019.csv")) %>% 
   clean_names() %>% 
   rename(value_2019 = x2019,
          hs_code = commodity_code,
          hs_name = commodity_name_1,
          isscaap_name = commodity_name_2,
          flag = s) %>% 
   mutate(flag = case_when(flag == "E" ~ "estimate",
                           TRUE ~ "reported")) %>% 
   select(-unit_name)


saveRDS(object = seafood_trade_volume_2019, file = here("data", "tidy_data", "seafood_trade_volume_product_weight_by_partner_2019.rds"))

```

CN-8 trade codes

```{r}

raw_codes <- read_csv(here("data", "raw_data", "conversion_factors_cn_8_2001_to_2021.csv")) %>% 
   clean_names() %>% 
   filter(year == 2019) %>% 
   mutate(cn_8 = gsub(" ", "", cn_8))



saveRDS(object = raw_codes, here("data", "tidy_data", "cn_8_codes.rds"))



#old raw codes tidy ofr UN trade codes

# names(raw_codes) <- "col"
# 
# 
# codes <- 
#    separate(data = raw_codes, col = col,  into=letters[1:10], sep = "\"") %>% 
#    select(a, b) %>% 
#    mutate(a = gsub(pattern = ",", replacement = "", x=a)) %>% 
#    rename(fao_code = a,
#           commodity_name = b)
# 
# saveRDS(object = codes, file = here("data", "tidy_data", "fao_product_codes.rds"))
# 

```



Seafood trade data from Jessica Gephart for 2018
```{r}

read_csv(here("data", "raw_data", "model0_2018_AUS_JAG.csv")) %>% 
   saveRDS(here("data", "tidy_data", "gephart_australian_imports.rds"))


```


Full food balance data - tidied to Australia only to speed up read in in other scripts
```{r}

fbs <- readRDS(here("data", "raw_data", "FoodBalanceSheets_E_All_Data_(Normalized).rds"))

fbs_aus <- fbs %>% clean_names() %>% filter(area == "Australia")

saveRDS(object = fbs_aus, file = here("data", "tidy_data", "food_balance_aus_1961-2013.rds"))

```


World bank ESG indicators for government effectiveness - taken from World Governance indicators
```{r}

wb_ESG <- read_csv(here("data", "raw_data", "world_bank_ESG_data.csv")) 

gov_eff <- wb_ESG %>% 
   clean_names() %>% 
   filter(series_name == "Government Effectiveness: Estimate") %>% 
   pivot_longer(cols = -c(country_name, country_code, series_name, series_code), names_to = "year", values_to =  "gov_effect") %>% 
   mutate(year = substring(text=year, first = 2, last = 5) %>% 
             as.numeric,
          gov_effect = gov_effect %>% as.numeric) %>% 
   drop_na(gov_effect) %>% 
   select(country_name, country_code, year, gov_effect) %>% 
   rename(iso3c = country_code)

saveRDS(object = gov_eff, file = here("data", "tidy_data", "government_effectiveness.rds"))

```



Global slavery index
```{r}

read_csv(file = here("data", "raw_data", "gsi.csv")) %>% 
   saveRDS(file = here("data", "tidy_data", "gsi.rds"))

```
Environmental Performance Index
```{r}

epi_2018 <- read_csv(here("data", "raw_data", "epi_2018.csv"))
epi_2020 <- read_csv(here("data", "raw_data", "epi_2020.csv"))

ecosystem_viability <- epi_2020 %>% select(iso, country, ECO.new) %>% 
   rename(iso3c = iso)

saveRDS(object = ecosystem_viability, file = here("data", "tidy_data", "ecosystem_vaibility.rds"))


```


Spatialised Australian landings raster
```{r}


wgs_84 <- "EPSG:4236"

aus_spatial <- readRDS(here("data/raw_data/aus_landings.rds")) %>% 
   group_by(Cell, LatCentre, LonCentre) %>% 
   summarise(reported = sum(Reported)) %>% 
   st_as_sf(coords = c("LonCentre", "LatCentre"))

raster_0.5 <- raster(crs=wgs_84, res=0.5)
points <- st_coordinates(aus_spatial$geometry)


aus_fishing <- rasterize(points, raster_0.5, field = aus_spatial$reported, fun = sum)

plot(aus_fishing)
writeRaster(x=aus_fishing, filename = here("data/spatial/aus_industrial_fishing.tif"), overwrite=TRUE)

```

