---
title: "tidy_data"
output: html_document
---

#libraries
```{r}

library(tidyverse)
library(janitor)
library(here)
library(countrycode)
library(vroom)
library(raster)
library(sf)

source(here("src/directories.R"))

select <- dplyr::select

```

Capture fisheries production data
```{r}
#Sort the symbols in values out
## "..." = data unavailable (here taken as zero)
## " " = data not separately available (here taken as zero)
## "-" = nil or zero (make zero)
## "0 0" = more than zero but less than half the unit used (taken to be 0.25)
## F = FAO estimate from available sources of information

col_seq <- seq(from = 8, to = 148, by =2)

remove_flags <- read_csv(here("data/raw_data/global_capture_production_1950_2020.csv")) |> 
   clean_names() |> 
   select(-unit) |> 
   filter(country_name == "Australia") |> 
   rename(country = country_name,
          species = asfis_species_name,
          isscaap_group = asfis_species_name_1,
          fishing_area = fao_major_fishing_area_name,
          fishing_area_code = fao_major_fishing_area_code, 
          unit = unit_name ) |> 
   select(-col_seq)
   
isolate_flags <- read_csv(here("data/raw_data/global_capture_production_1950_2020.csv")) |> 
   clean_names() |> 
   select(-unit) |> 
   filter(country_name == "Australia") |> 
   rename(country = country_name,
          species = asfis_species_name,
          isscaap_group = asfis_species_name_1,
          fishing_area = fao_major_fishing_area_name,
          fishing_area_code = fao_major_fishing_area_code, 
          unit = unit_name ) |> 
   select(col_seq)

names(isolate_flags) <- paste0("f_", seq(1950, 2020))

capture_prod <- 
   remove_flags |> 
   pivot_longer(cols = -c(1:6), names_to = "year", values_to = "value") |> 
   mutate(year = gsub(pattern = "x", replacement = "", year) |> as.numeric())


saveRDS(object = capture_prod, file = here("data", "tidy_data", "capture_production_tidy.rds"))


```


Aquaculture production data
```{r}

col_seq <- seq(from = 9, to = 149, by =2)

remove_flags_aqua <- 
   read_csv(here("data/raw_data/global_aquaculture_production_1950_2020.csv")) |> 
   clean_names() |> 
   select(-unit) |> 
   filter(country_name == "Australia") |> 
   rename(country = country_name,
          species = asfis_species_name,
          isscaap_group = asfis_species_name_1,
          fishing_area = fao_major_fishing_area_name,
          fishing_area_code = fao_major_fishing_area_code, 
          environment = environment_name,
          unit = unit_name ) |> 
   select(-col_seq)

aqua_prod <- 
   remove_flags |> 
   pivot_longer(cols = -c(1:6), names_to = "year", values_to = "value") |> 
   mutate(year = gsub(pattern = "x", replacement = "", year) |> as.numeric())


saveRDS(object = aqua_prod, file = here("data", "tidy_data", "aquaculture_production_tidy.rds"))


```


Fish food balance data
```{r}

(fish_fbs_raw <- read_csv(here("data", "raw_data", "fish_FBS_1961-2017.csv")) %>% 
   clean_names()
)

#remove every other column from row 6

col_seq <- seq(from = 6, to = 118, by=2)

fish_fbs_raw <- fish_fbs_raw[, -col_seq]

(fish_fbs <- fish_fbs_raw %>% 
      pivot_longer(cols = -c(country_name, faostat_group_name, element_name, unit_name), names_to = "year", values_to = "value") %>%
      mutate(year = gsub("x", "", year))
)


saveRDS(object = fish_fbs, file = here("data", "tidy_data", "fish_fbs_tidy.rds"))

```


Fish trade data - volume, temporal
```{r}

col_seq <- seq(from = 7, to = 97, by=2) #remove every other column from the 7th

seafood_trade_volume <- read_csv(here("data", "raw_data", "fish_trade_volume_1976-2020.csv")) %>% 
   clean_names() %>% 
   select(-col_seq) %>% 
   filter(!reporting_country_name %in% c("Totals - Tonnes – net product weight", "FAO. 2021. Fishery and Aquaculture Statistics. Global Fish Trade - All partners aggregated 1976-2019 (FishstatJ). In: FAO Fisheries and Aquaculture Division [online]. Rome. Updated 2021. www.fao.org/fishery/statistics/software/fishstatj/en"))  |> 
   rename(hs_code = commodity_code,
          isscaap_name = commodity_name_1) |> 
   pivot_longer(names_to = "year", values_to = "value", cols=-c(1:6)) %>% 
   mutate(year = gsub("x", "", year) %>% 
             as.numeric)


saveRDS(object = seafood_trade_volume, file = here("data", "tidy_data", "seafood_trade_volume_product_weight_1976_2020.rds"))
   


```


Fish trade data - value, temporal
```{r}

(seafood_value <- read_csv(here("data", "raw_data", "fish_trade_value_1976-2019.csv")) %>% 
   clean_names() %>% 
    select(-seq(from = 7, to = 93, by =2)) %>% 
 pivot_longer(cols=-c(1:5), names_to = "year", values_to = "value") %>% 
    mutate(year = gsub("x", "", year) %>% 
              as.numeric,
           value = value*1000,
           unit = "USD",
           unit_name = "Value")
)

saveRDS(object = seafood_value, file = here("data", "tidy_data", "seafood_trade_value_product_weight_1976-2019.rds"))

```



Fish trade - volume by trade partner in 2019
```{r}

seafood_trade_volume_2019 <- read_csv(here("data", "raw_data", "fish_trade_volume_by_partner_2019.csv")) %>% 
   clean_names() %>% 
   rename(value_2019 = x2019,
          hs_code = commodity_code,
          hs_name = commodity_name_1,
          isscaap_name = commodity_name_2,
          flag = s) %>% 
   mutate(flag = case_when(flag == "E" ~ "estimate",
                           TRUE ~ "reported")) %>% 
   select(-unit_name)


saveRDS(object = seafood_trade_volume_2019, file = here("data", "tidy_data", "seafood_trade_volume_product_weight_by_partner_2019.rds"))

```

CN-8 trade codes

```{r}

raw_codes <- read_csv(here("data", "raw_data", "conversion_factors_cn_8_2001_to_2021.csv")) %>% 
   clean_names() %>% 
   filter(year == 2019) %>% 
   mutate(cn_8 = gsub(" ", "", cn_8))



saveRDS(object = raw_codes, here("data", "tidy_data", "cn_8_codes.rds"))



#old raw codes tidy ofr UN trade codes

# names(raw_codes) <- "col"
# 
# 
# codes <- 
#    separate(data = raw_codes, col = col,  into=letters[1:10], sep = "\"") %>% 
#    select(a, b) %>% 
#    mutate(a = gsub(pattern = ",", replacement = "", x=a)) %>% 
#    rename(fao_code = a,
#           commodity_name = b)
# 
# saveRDS(object = codes, file = here("data", "tidy_data", "fao_product_codes.rds"))
# 

```



Seafood trade data from Jessica Gephart for 2018
```{r}

read_csv(here("data", "raw_data", "model0_2018_AUS_JAG.csv")) %>% 
   saveRDS(here("data", "tidy_data", "gephart_australian_imports.rds"))


```


Full food balance data - tidied to Australia only to speed up read in in other scripts
```{r}

fbs <- readRDS(here("data", "raw_data", "FoodBalanceSheets_E_All_Data_(Normalized).rds"))

fbs_aus <- fbs %>% clean_names() %>% filter(area == "Australia")

saveRDS(object = fbs_aus, file = here("data", "tidy_data", "food_balance_aus_1961-2013.rds"))

```


World bank ESG indicators for government effectiveness - taken from World Governance indicators
```{r}

wb_ESG <- read_csv(here("data", "raw_data", "world_bank_ESG_data.csv")) 

gov_eff <- wb_ESG %>% 
   clean_names() %>% 
   filter(series_name == "Government Effectiveness: Estimate") %>% 
   pivot_longer(cols = -c(country_name, country_code, series_name, series_code), names_to = "year", values_to =  "gov_effect") %>% 
   mutate(year = substring(text=year, first = 2, last = 5) %>% 
             as.numeric,
          gov_effect = gov_effect %>% as.numeric) %>% 
   drop_na(gov_effect) %>% 
   select(country_name, country_code, year, gov_effect) %>% 
   rename(iso3c = country_code)

saveRDS(object = gov_eff, file = here("data", "tidy_data", "government_effectiveness.rds"))

```



Global slavery index
```{r}

read_csv(file = here("data", "raw_data", "gsi_vuln.csv")) |> 
   mutate(iso3c = countrycode(sourcevar = country, origin = "country.name", destination = "iso3c", warn=TRUE)) |> 
   saveRDS(file = here("data", "tidy_data", "gsi.rds"))

```

Environmental Performance Index
```{r}

epi_2018 <- read_csv(here("data", "raw_data", "epi_2018.csv"))
epi_2020 <- read_csv(here("data", "raw_data", "epi_2020.csv"))

ecosystem_vitality <- epi_2020 %>% select(iso, country, ECO.new) %>% 
   rename(iso3c = iso)

saveRDS(object = ecosystem_vitality, file = here("data", "tidy_data", "ecosystem_vitality.rds"))

epi <- epi_2020 %>% select(iso, country, EPI.new) %>% 
   rename(iso3c = iso)

saveRDS(object = epi, file = here("data", "tidy_data", "environmental_performance_index.rds"))

```


EIU Illicit trade index 
```{r}

eiu_illicit_trade <- read_csv(file.path(rdsi_raw_data_dir, "eiu-illicit-trade-index/eiu-illicit-trade-index.csv")) |> 
   rename(illicit_score = epi_score) |> 
   mutate(iso3c = countrycode(sourcevar = country_name, origin = "country.name", destination = "iso3c", warn = TRUE))

#save in project
saveRDS(object = eiu_illicit_trade, file = here("data/tidy_data/eiu-illicit-trade.rds"))

```



Spatialised Australian landings raster
```{r}


wgs_84 <- "EPSG:4236"

aus_spatial <- readRDS(here("data/raw_data/aus_landings.rds")) %>% 
   group_by(Cell, LatCentre, LonCentre) %>% 
   summarise(reported = sum(Reported)) %>% 
   st_as_sf(coords = c("LonCentre", "LatCentre"))

raster_0.5 <- raster(crs=wgs_84, res=0.5)
points <- st_coordinates(aus_spatial$geometry)


aus_fishing <- rasterize(points, raster_0.5, field = aus_spatial$reported, fun = sum)

plot(aus_fishing)
writeRaster(x=aus_fishing, filename = here("data/spatial/aus_industrial_fishing.tif"), overwrite=TRUE)

```

Clean and join sentiment ratings from Scott and Gabriel

```{r}


(scott_sentiment <- read_csv(here("data/raw_data/scott_sentiment.csv")) %>% 
    clean_names() %>% 
    arrange(state) %>% 
    dplyr::select(-c(system, sentiment_2, sentiment_1_raw, metadata))
)



(gabriel_sentiment <-  read_csv(here("data/raw_data/gabriel_sentiment.csv")) %>% 
    clean_names() %>% 
    arrange(state) %>% 
      dplyr::select(-c(system, sentiment_1, sentiment_2_numerical, metadata))
)


(combined_sentiment <- scott_sentiment %>% 
   left_join(gabriel_sentiment, by=c("state", "headline")) %>% 
   mutate(sentiment_1 = if_else(is.na(sentiment_1), true = "NA", sentiment_1),
           sentiment_2 = if_else(is.na(sentiment_2), true = "NA", sentiment_2)) %>% 
      mutate(agreement_1_2 = if_else(sentiment_1==sentiment_2, true = TRUE, false = FALSE))
)


write_csv(x = combined_sentiment, file = here("data/raw_data/combined_sentiment.csv")) %>% 
  saveRDS(file = here("data/tidy_data/combined_sentiment.rds"))

```


Convert agreed sentiment to rds
```{r}

read_csv(here("data/raw_data/combined_sentiment_agreed.csv")) %>% 
  saveRDS(file = here("data/tidy_data/combined_sentiment_agreed.rds"))


```


Tidy Aus state level aquaculture data
```{r}

aus_state_aqua <- read_csv(here("data/raw_data/aus_aquaculture_production.csv")) %>%
  pivot_longer(names_to = "year", values_to = "quantity", -c(state, species)) %>% 
  mutate(quantity = gsub(",", "", quantity)) %>% 
  mutate(year = as.integer(year),
         quantity = as.double(quantity))


saveRDS(aus_state_aqua, here("data/tidy_data/state_aquaculture_production_tidy.rds"))


```

Import Aus stock assessment data

```{r}

stock_summary <- read_csv(file.path(rdsi_raw_data_dir, "fishery-stock-status/aus_2012_2020.csv")) |> 
   group_by(year) |> 
   nest() |> 
   mutate(sum = map(data, ~(sum(.$freq)))) |> 
   unnest(c(data, sum)) |> 
   ungroup() |> 
   mutate(prop = freq/sum)

write_csv(x = stock_summary, file = here("data/tidy_data/fishery_stock_status.csv"))
```

